---
title: "Pieris virginiensis"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.height = 3)
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(parallel))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(ggmap))
```

```{r data-entry}
genus_name <- "Pieris"
species_name <- "virginiensis"
long_limits <- c(-99, -45)
lat_limits <- c(15, 70)

# Path to folder that contains retrieval script, get_clean_obs_function.R
retrieval_path <- "../insect_migration/scripts/"
retrieval_name <- "get_clean_obs_function.R"
```

```{r check-values, results = "asis", warning = FALSE}
# Keep track of whether a report can be assembled after the setup-check
continue_knitting <- TRUE

# Seems excessive...
if (file.exists("scripts/global_values.R")) {
  source(file = "scripts/global_values.R")
  globals <- global_values()
} else {
  cat("\n## Missing global values script\n")
  cat("Make sure scripts/global_values.R exists")
  continue_knitting <- FALSE
}

# Make sure retrieval_path has trailing /
if (substr(x = retrieval_path, start = nchar(retrieval_path), stop = nchar(retrieval_path)) != "/") {
  retrieval_path <- paste0(retrieval_path, "/")
}

# Check to make sure get_clean_obs_function exists
if (!file.exists(paste0(retrieval_path, retrieval_name))) {
  cat("\n## Missing required function `get_clean_obs_function`.\n")
  cat("Please check `retrieval_path` setting in `data-entry` chunk of Rmarkdown file. The function can be obtained from GitHub via `git clone https://github.com/keatonwilson/insect_migration.git`\n")
  continue_knitting <- FALSE
}

if ("FIXME" %in% c(species_name, genus_name)) {
  cat("\n## `FIXME` values need to be changed in `data-entry` chunk of Rmarkdown file.\n")
  continue_knitting <- FALSE
}

if (any(is.na(c(long_limits, lat_limits)))) {
  cat("\n## `NA` values need to be changed in `data-entry` chunk of Rmarkdown file.\n")
  continue_knitting <- FALSE
}

required_pkgs <- c("tidyverse", "spocc", "mapr")
missing_pkgs <- c()
for (pkg in required_pkgs) {
  if (!suppressPackageStartupMessages(library(package = pkg, logical.return = TRUE, character.only = TRUE))) {
    missing_pkgs <- c(missing_pkgs, pkg)
  }
}
if (length(missing_pkgs) > 0) { 
  cat("\n## Missing one or more required libraries.\n")
  cat("Please install the following package(s): ", paste(missing_pkgs, collapse = ", "), ".\n", sep = "")
  continue_knitting <- FALSE
}
```

```{r status-check-setup}
# If there are problems at this point, don't bother evaluating the rest of the 
# document
if (!continue_knitting) {
  knitr::knit_exit()
}
```

# Report for _`r paste(genus_name, species_name, sep = " ")`_

```{r download-data, include = FALSE}
# Use include = FALSE to suppress warning messages
source(file = paste0(retrieval_path, retrieval_name))

data_file <- paste0("data/", genus_name, "-", species_name, "-spocc.csv")

if (file.exists(data_file)) {
  observations <- read_csv(file = data_file)
} else {
  observations <- get_clean_obs(genus = genus_name, 
                                species = species_name, 
                                lonlim = long_limits, 
                                latlim = lat_limits)
# UPDATE: This error is likely caused by local network issues causing long 
# address resolution leading to timeouts
# TODO: need to test this with get_clean_obs_function; the above throws:
# Quitting from lines 87-92 (Pieris-virginiensis-emergence.Rmd) 
#  Error: object 'longitude' not found
# the error happens somewhere around:
# mutate_impl(.data, dots, caller_env()) 
# mutate.tbl_df(., longitude = as.numeric(longitude), latitude = as.numeric(latitude)) 
# mutate(., longitude = as.numeric(longitude), latitude = as.numeric(latitude)) 

}

if (nrow(observations) > 0) {
  write_csv(x = observations, path = data_file)
} else {
  cat("\n## No records returned for _", genus_name, " ", species_name, "_.\n")
  cat("Check to be sure names are spelled correctly and iNat and GBIF servers are responsive")
  continue_knitting <- FALSE
}
```

```{r status-check-data}
# If there are problems at this point, don't bother evaluating the rest of the 
# document
if (!continue_knitting) {
  knitr::knit_exit()
}
```

Retrieval from iNaturalist and GBIF returned `r nrow(observations)` observations.

```{r data-cleaning}
# Restrict to the days and years of interest
observations <- observations %>%
  filter(!is.na(date),
         yday(x = date) >= globals$min_julian,
         yday(x = date) <= globals$max_julian,
         year(x = date) >= globals$min_year,
         year(x = date) <= globals$max_year)

if (nrow(observations) == 0) {
  cat("\n## Zero observations remain after date filtering.\n")
  cat("Current filters only include ", globals$min_year, "-", globals$max_year)
  cat(" and julian days ", globals$min_julian, "-", globals$max_julian, ".\n")
  continue_knitting <- FALSE
}
```

```{r status-check-cleaning}
# If there are problems at this point, don't bother evaluating the rest of the 
# document
if (!continue_knitting) {
  knitr::knit_exit()
}
```

```{r meet-minimums}
# Calculate which years meet the minimum number of required samples
year_counts <- observations %>%  
  group_by(year = year(date)) %>%
  summarize(obs_count = n()) %>%
  filter(obs_count >= globals$minimum_required)

# Filter obserations to only include those years with enough samples
observations <- observations %>%
  filter(year(date) %in% year_counts$year)

if (nrow(observations) == 0) {
  cat("\n## Zero observations remain after removing years with too few observations\n")
  cat("Current filter requires at least ", globals$minimum_required, 
      " observations per year", ".\n")
  continue_knitting <- FALSE
}
```

```{r status-check-minimums}
# If there are problems at this point, don't bother evaluating the rest of the 
# document
if (!continue_knitting) {
  knitr::knit_exit()
}
```


Distribution of _`r paste(genus_name, species_name, sep = " ")`_ data for `r min(year(observations$date))` - `r max(year(observations$date))` (total of `r nrow(observations)` observations following date and sample size filtering):

```{r plot-data}
# Plot a map of observations post-filtering

# Only download map files if we haven't already
map_file <- data_file <- paste0("data/maps/", genus_name, "-", species_name, "-map.RData")

if (file.exists(map_file)) {
  load(file = map_file)
} else {
  # Get a map image
  # Have to provide min/max for lat and long, otherwise will assume Google map,
  # regardless of value sent to source parameter
  map_bounds <- c(floor(min(observations$longitude)),
                floor(min(observations$latitude)),
                ceiling(max(observations$longitude)),
                ceiling(max(observations$latitude)))

  obs_map <- get_map(location = map_bounds,
                     source = "stamen",
                     maptype = "terrain")
  save(obs_map, file = map_file)
}

map_plot <- ggmap(ggmap = obs_map) +
  geom_point(data = observations,
             mapping = aes(x = longitude, y = latitude),
             size = 1.0) +
  theme_bw() +
  ylab(label = "Latitude") +
  xlab(label = "Longitude")

# Plot observations by year
obs_plot <- ggplot(data = observations,
                   mapping = aes(x = year(date),
                                 y = yday(date),
                                 color = latitude)) +
  geom_point() +
  xlab(label = "Year") +
  scale_color_gradient(name = "Latitude") +
  ylab(label = "Day") +
  theme_bw() +
  theme(legend.position = "none", #c(0.12, 0.8),
        legend.background = element_rect(size = 1.5),
        legend.title = element_text(size = 6),
        legend.text = element_text(size = 6),
        legend.key.size = unit(x = 0.7, units = "line"))

# Two-panel plot of map and time
suppressMessages(expr = grid.arrange(map_plot, 
                                     obs_plot, 
                                     nrow = 1))
```
Note in left-hand plot, darker points are from lower latitudes, while lighter points are from higher latitudes.

# Test for change in emergences
## Bootstrapping
```{r bootstrap-linear-regression}
# A data frame we'll use as template for storing bootstrapped data
bootstrapped_df <- data.frame(year = rep(year_counts$year, 
                                         times = globals$bs_sample_size),
                              yday = NA)


# Set up parallel processing
num_cores <- detectCores() - 1

# Initiate cluster
the_cluster <- makeCluster(num_cores)

# Need to explicitly load packages in cluster; invisible the only way to prevent
# output created by loading tidyverse on cores
invisible(x = clusterEvalQ(cl = the_cluster, { library(tidyverse); library(lubridate) }))

bs_results <- parLapply(cl = the_cluster,
                        X = c(1:globals$num_bs_reps),
                        # X is passed into the i argument below
                        fun = function(i, bs_df, sample_df, sample_size) {
                          # Fill in data bootstrapped for each year, sampling
                          # *only* from sample_vec, and restricting sample 
                          # size of each year to sample_size
                          for (yr in unique(bs_df$year)) {
                            sample_vec <- yday(sample_df$date)[year(sample_df$date) == yr]
                            bs_df$yday[bs_df$year == yr] <- sample(x = sample_vec,
                                                                   size = sample_size,
                                                                   replace = FALSE)
                          }
                          # return(i)

                          bs_mins <- bs_df %>%
                            group_by(year) %>%
                            summarise(early = min(yday))

                          bs_lm <- lm(early ~ year, data = bs_mins)
                          coeff_est <- summary(bs_lm)$coefficients[2, 1]
                          intercept <- summary(bs_lm)$coefficients[1, 1]
                          p_value <- summary(bs_lm)$coefficients[2, 4]
                          return(data.frame(bs_rep = i,
                                            estimate = coeff_est,
                                            intercept = intercept,
                                            p_value = p_value))
                        },
                        bs_df = bootstrapped_df,
                        sample_df = observations,
                        sample_size = globals$bs_sample_size)

stopCluster(the_cluster)
# Need to convert vanilla list into data frame
bs_results <- do.call(rbind, bs_results)
```

### Linear regression
```{r lm-reporting}
mean_effect_size <- mean(bs_results$estimate)
direction <- ifelse(mean_effect_size < 0,
                    yes = "earlier",
                    no = "later")
p_value <- ifelse(mean_effect_size < 0,
                  yes = sum(bs_results$estimate > 0),
                  no = sum(bs_results$estimate < 0)) / globals$num_bs_reps

p_value_text <- paste0("= ", round(x = p_value, digits = 3))
if (p_value == 0) {
  p_value_text <- paste0("< ", round(x = 1/globals$num_bs_reps, digits = 3))
}
```

In `r globals$num_bs_reps` bootstrap replicates, the mean change in emergence times was `r round(x = abs(mean_effect_size), digits = 3)` days `r direction` per year (p `r p_value_text`).

```{r plot-bootstrap-lm}
# Plot the bootstrap results

# Plot the coefficient estimates
plot_title <- paste0("Effect size from ",
                        globals$num_bs_reps,
                        " bootstraps")

estimates_plot <- ggplot(data = bs_results, 
                            mapping = aes(x = estimate)) +
  geom_histogram() +
  geom_vline(xintercept = 0.0, color = "red") +
  xlab(label = "Effect size (days/year)") +
  ylab(label = "Count") +
  ggtitle(label = plot_title) + 
  theme_bw() +
  theme(plot.title = element_text(size = 10))

# Plot the results of bootstrapping with early dates
emergences <- observations %>%
  group_by(year = year(date)) %>%
  summarise(early = min(yday(date), na.rm = TRUE),
            n_obs = n())

emergences_title <- paste0("Earliest observations, ", 
                     min(year(observations$date)), " - ", 
                     max(year(observations$date)))

emergences_subtitle <- paste0("Bootstrapped trendlines, ",
                            globals$num_bs_reps,
                            " replicates")
emergences_plot <- ggplot(data = emergences,
                        mapping = aes(x = year, y = early)) +
  geom_point() +
  geom_abline(data = bs_results,
              alpha = 5/globals$num_bs_reps,
              mapping = aes(slope = bs_results$estimate, 
                            intercept = bs_results$intercept)) +
  xlab(label = "Year") +
  ylab(label = "Day") +
  ggtitle(emergences_title,
          subtitle = emergences_subtitle) + 
  theme_bw() +
  theme(plot.title = element_text(size = 10),
        plot.subtitle = element_text(size = 8))

suppressMessages(expr = grid.arrange(estimates_plot, 
                                     emergences_plot, 
                                     nrow = 1))
```

### Linear spline
## Weighted least-squares regression
```{r wls-linear}
# Pull out eariest date for each year, as well as number of observations for 
# that year; the number of observations will be used as the weight for that year
earliest <- observations %>%
  group_by(year = year(date)) %>%
  summarize(early = min(x = yday(date)),
            n_obs = n())
earliest <- as.data.frame(earliest)

# Use number of observations as weight for each year
wls_model <- lm(early ~ year, data = earliest, weights = n_obs)
wls_summary <- summary(wls_model)
wls_estimate <- wls_summary$coefficients[2, 1]

# Only want to show 3 digits without rounding to 0
wls_p_value <- formatC(x = wls_summary$coefficients[2, 4],
                       digits = 3,
                       format = "e")

wls_direction <- ifelse(wls_estimate < 0,
                        yes = "earlier",
                        no = "later")

```

### Linear regression
In weighted least-squares regression, using a simple linear model with observations weighted by the number of observations in that year, the effect size was `r round(x = abs(wls_estimate), digits = 3)` days `r wls_direction` per year (p = `r wls_p_value`).

### Linear spline
```{r wls-spline}
# Testing smoothing spline
# TODO: probably want to find a way to do a linear spline
spline_model <- smooth.spline(x = earliest$year,
                              y = earliest$early,
                              w = earliest$n_obs,
                             nknots = 4) # <=3 produces errors
```