---
title: "Pieris virginiensis progress report 03"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
geometry: margin=0.75in
params: 
    genus_name: "Pieris"
    species_name: "virginiensis"
    min_lon: -99
    max_lon: -45
    min_lat: 15
    max_lat: 70
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.height = 3)
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(parallel))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(ggmap))
```


```{r data-entry}
genus_name <- params$genus_name
species_name <- params$species_name
long_limits <- c(params$min_lon, params$max_lon)
lat_limits <- c(params$min_lat, params$max_lat)

# Path to folder that contains retrieval script, get_clean_obs_function.R
retrieval_path <- "../insect_migration/scripts/"
retrieval_name <- "get_clean_obs_function.R"
```

```{r check-values, results = "asis", warning = FALSE}
# Keep track of whether a report can be assembled after the setup-check
continue_knitting <- TRUE

# Seems excessive...
if (file.exists("scripts/global_values.R")) {
  source(file = "scripts/global_values.R")
  globals <- global_values()
} else {
  cat("\n## Missing global values script\n")
  cat("Make sure scripts/global_values.R exists")
  continue_knitting <- FALSE
}

# Make sure retrieval_path has trailing /
if (substr(x = retrieval_path, start = nchar(retrieval_path), stop = nchar(retrieval_path)) != "/") {
  retrieval_path <- paste0(retrieval_path, "/")
}

# Check to make sure get_clean_obs_function exists
if (!file.exists(paste0(retrieval_path, retrieval_name))) {
  cat("\n\n## Missing required function `get_clean_obs_function`\n")
  cat("Please check `retrieval_path` setting in `data-entry` chunk of Rmarkdown file. The function can be obtained from GitHub via `git clone https://github.com/keatonwilson/insect_migration.git`\n")
  continue_knitting <- FALSE
}

if ("FIXME" %in% c(species_name, genus_name)) {
  cat("\n## `FIXME` values need to be changed in `data-entry` chunk of Rmarkdown file.\n")
  continue_knitting <- FALSE
}

if (any(is.na(c(long_limits, lat_limits)))) {
  cat("\n## `NA` values need to be changed in `data-entry` chunk of Rmarkdown file.\n")
  continue_knitting <- FALSE
}

required_pkgs <- c("tidyverse", "spocc", "mapr")
missing_pkgs <- c()
for (pkg in required_pkgs) {
  if (!suppressPackageStartupMessages(library(package = pkg, logical.return = TRUE, character.only = TRUE))) {
    missing_pkgs <- c(missing_pkgs, pkg)
  }
}
if (length(missing_pkgs) > 0) { 
  cat("\n## Missing one or more required libraries.\n")
  cat("Please install the following package(s): ", paste(missing_pkgs, collapse = ", "), ".\n", sep = "")
  continue_knitting <- FALSE
}
```

```{r status-check-setup}
# If there are problems at this point, don't bother evaluating the rest of the 
# document
if (!continue_knitting) {
  knitr::knit_exit()
}
```

## Data for _`r paste(genus_name, species_name, sep = " ")`_

```{r download-data, include = FALSE}
# Use include = FALSE to suppress warning messages
source(file = paste0(retrieval_path, retrieval_name))

data_file <- paste0("data/", genus_name, "-", species_name, "-spocc.csv")

# Load in data, if they exist on disk. If not, download via spocc
if (file.exists(data_file)) {
  observations <- read_csv(file = data_file)
} else {
  observations <- get_clean_obs(genus = genus_name, 
                                species = species_name, 
                                lonlim = long_limits, 
                                latlim = lat_limits)
}

# Check for zero observations retrieved
if (nrow(observations) > 0) {
  write_csv(x = observations, path = data_file)
  check_name <- FALSE
  if (nrow(observations[observations$prov == "gbif", ]) == 0) {
    cat("\n### Note: zero records were returned from GBIF\n")
    check_name <- TRUE
  }
  if (nrow(observations[observations$prov == "inat", ]) == 0) {
    cat("\n### Note: zero records were returned from iNaturalist\n")
    check_name <- TRUE
  }
  if (check_name) {
    cat("Please check to see if _", genus_name, " ", species_name, 
        "_ is the name used for the data provider.\n")
  }
} else {
  cat("\n## No records returned for _", genus_name, " ", species_name, "_.\n")
  cat("Check to be sure names are spelled correctly and iNat and GBIF servers are responsive")
  continue_knitting <- FALSE
}
```

```{r status-check-data}
# If there are problems at this point, don't bother evaluating the rest of the 
# document
if (!continue_knitting) {
  knitr::knit_exit()
}
```

Retrieval from iNaturalist and GBIF returned `r nrow(observations)` observations.

```{r data-cleaning}
# Restrict to the days and years of interest
observations <- observations %>%
  filter(!is.na(date),
         yday(x = date) >= globals$min_julian,
         yday(x = date) <= globals$max_julian,
         year(x = date) >= globals$min_year,
         year(x = date) <= globals$max_year)

if (nrow(observations) == 0) {
  cat("\n## Zero observations remain after date filtering.\n")
  cat("Current filters only include ", globals$min_year, "-", globals$max_year)
  cat(" and julian days ", globals$min_julian, "-", globals$max_julian, ".\n")
  continue_knitting <- FALSE
}
```

```{r status-check-cleaning}
# If there are problems at this point, don't bother evaluating the rest of the 
# document
if (!continue_knitting) {
  knitr::knit_exit()
}
```

```{r meet-minimums}
# Calculate which years meet the minimum number of required samples
year_counts <- observations %>%  
  group_by(year = year(date)) %>%
  summarize(obs_count = n()) %>%
  filter(obs_count >= globals$minimum_required)

# Filter obserations to only include those years with enough samples
observations <- observations %>%
  filter(year(date) %in% year_counts$year)

if (nrow(observations) == 0) {
  cat("\n## Zero observations remain after removing years with too few observations\n")
  cat("Current filter requires at least ", globals$minimum_required, 
      " observations per year", ".\n")
  continue_knitting <- FALSE
}
```

```{r status-check-minimums}
# If there are problems at this point, don't bother evaluating the rest of the 
# document
if (!continue_knitting) {
  knitr::knit_exit()
}
```


Distribution of _`r paste(genus_name, species_name, sep = " ")`_ data for `r min(year(observations$date))` - `r max(year(observations$date))` (total of `r nrow(observations)` observations following date and sample size filtering):

```{r plot-data, include = FALSE, fig.height = 2.5}
# include = FALSE to prevent ggmap::get_map messages from showing up in output
# Plot a map of observations post-filtering

# Only download map files if we haven't already
map_file <- data_file <- paste0("data/maps/", genus_name, "-", species_name, "-map.RData")

if (file.exists(map_file)) {
  load(file = map_file)
} else {
  # Get a map image
  # Have to provide min/max for lat and long, otherwise will assume Google map,
  # regardless of value sent to source parameter
  map_bounds <- c(floor(min(observations$longitude)),
                floor(min(observations$latitude)),
                ceiling(max(observations$longitude)),
                ceiling(max(observations$latitude)))

  obs_map <- get_map(location = map_bounds,
                     source = "stamen",
                     maptype = "terrain")
  save(obs_map, file = map_file)
}

map_plot <- ggmap(ggmap = obs_map) +
  geom_point(data = observations,
             mapping = aes(x = longitude, y = latitude),
             size = 1.0) +
  theme_bw() +
  ylab(label = "Latitude") +
  xlab(label = "Longitude")

# Plot observations by year
obs_plot <- ggplot(data = observations,
                   mapping = aes(x = year(date),
                                 y = yday(date),
                                 color = latitude)) +
  geom_point() +
  xlab(label = "Year") +
  scale_color_gradient(name = "Latitude") +
  ylab(label = "Day") +
  theme_bw() +
  theme(legend.position = "none", #c(0.12, 0.8),
        legend.background = element_rect(size = 1.5),
        legend.title = element_text(size = 6),
        legend.text = element_text(size = 6),
        legend.key.size = unit(x = 0.7, units = "line"))

```

```{r draw-observation-plots}
# Two-panel plot of map and time
suppressMessages(expr = grid.arrange(map_plot, 
                                     obs_plot, 
                                     nrow = 1))
```

Note in left-hand plot, **darker** points are from lower latitudes, while **lighter** points are from higher latitudes.

# Test for change in emergence

Starting with simple linear regression, not considering variation in sampling across years (but should probably use weighted least-squares regression for this).

```{r flights-latitude}
obs <- observations %>%
  mutate(yday = yday(date),
         year = year(date)) %>%
  select(latitude, yday, year)

lat_model <- lm(yday ~ year + latitude + year*latitude, data = obs)
lat_model_coeffs <- summary(lat_model)$coefficients

lat_model_coeffs <- round(lat_model_coeffs, digits = 3)

model_intercept <- summary(lat_model)$coefficients[1, 1]
model_year_slope <- summary(lat_model)$coefficients[2, 1]

# TODO: look at ggPredict package for plotting this
# OR, create a series of prediction lines for different latitudes...
# To create regression lines for this plot, re-arrange the linear regression model
# yday = B0 + B1 * year + B2 * latitude + B3 + year * latitude
# yday = B0 + B2 * latitude + (B1 + B3 * latitude) * year
# where the sum of the first two terms is the intercept and (B1 + B3 * latitude)
# is the slope
lat_plot <- ggplot(data = obs, mapping = aes(x = year, y = yday, color = latitude)) +
  geom_point() +
  ylab(label = "Julian Day") +
  xlab(label = "Year") +
  scale_color_gradient() +
  theme_bw()

print(lat_plot)
```
