---
title: "Pieris virginiensis progress report 04"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
geometry: margin=0.75in
params: 
    min_lon: -99
    max_lon: -45
    min_lat: 15
    max_lat: 70
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.height = 3)
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(parallel))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(ggmap))
```


```{r data-entry}
# Will want to create two data frames: one with insect and one with host(s) 
species <- data.frame(genus = c("Pieris", "Cardamine", "Cardamine"),
                      species = c("virginiensis", "concatenata", "diphylla"),
                      category = c("insect", "host", "host"),
                      stringsAsFactors = FALSE)

long_limits <- c(params$min_lon, params$max_lon)
lat_limits <- c(params$min_lat, params$max_lat)

# Path to folder that contains retrieval script, get_clean_obs_function.R
retrieval_path <- "../insect_migration/scripts/"
retrieval_name <- "get_clean_obs_function.R"
```

```{r check-values, results = "asis", warning = FALSE}
# Keep track of whether a report can be assembled after the setup-check
continue_knitting <- TRUE

# Seems excessive...
if (file.exists("scripts/global_values.R")) {
  source(file = "scripts/global_values.R")
  globals <- global_values()
} else {
  cat("\n## Missing global values script\n")
  cat("Make sure scripts/global_values.R exists")
  continue_knitting <- FALSE
}

# Make sure retrieval_path has trailing /
if (substr(x = retrieval_path, start = nchar(retrieval_path), stop = nchar(retrieval_path)) != "/") {
  retrieval_path <- paste0(retrieval_path, "/")
}

# Check to make sure get_clean_obs_function exists
if (!file.exists(paste0(retrieval_path, retrieval_name))) {
  cat("\n\n## Missing required function `get_clean_obs_function`\n")
  cat("Please check `retrieval_path` setting in `data-entry` chunk of Rmarkdown file. The function can be obtained from GitHub via `git clone https://github.com/keatonwilson/insect_migration.git`\n")
  continue_knitting <- FALSE
}

if (any(is.na(c(long_limits, lat_limits)))) {
  cat("\n## `NA` values need to be changed in `data-entry` chunk of Rmarkdown file.\n")
  continue_knitting <- FALSE
}

required_pkgs <- c("tidyverse", "spocc", "mapr")
missing_pkgs <- c()
for (pkg in required_pkgs) {
  if (!suppressPackageStartupMessages(library(package = pkg, logical.return = TRUE, character.only = TRUE))) {
    missing_pkgs <- c(missing_pkgs, pkg)
  }
}
if (length(missing_pkgs) > 0) { 
  cat("\n## Missing one or more required libraries.\n")
  cat("Please install the following package(s): ", paste(missing_pkgs, collapse = ", "), ".\n", sep = "")
  continue_knitting <- FALSE
}
```

```{r status-check-setup}
# If there are problems at this point, don't bother evaluating the rest of the 
# document
if (!continue_knitting) {
  knitr::knit_exit()
}
```

```{r download-data, results = "asis", include = FALSE}
# Use include = FALSE to suppress warning messages
source(file = paste0(retrieval_path, retrieval_name))

all_obs <- NULL

for (i in 1:nrow(species)) {
  genus_name <- species$genus[i]
  species_name <- species$species[i]
  category <- species$category[i]
  
  data_file <- paste0("data/", genus_name, "-", species_name, "-spocc.csv")
  
  # Load in data, if they exist on disk. If not, download via spocc
  if (file.exists(data_file)) {
    observations <- read_csv(file = data_file)
  } else {
    observations <- get_clean_obs(genus = genus_name, 
                                  species = species_name, 
                                  lonlim = long_limits, 
                                  latlim = lat_limits)
    
    # Check for zero observations retrieved
    if (nrow(observations) > 0) {
      write_csv(x = observations, path = data_file)
      check_name <- FALSE
      if (nrow(observations[observations$prov == "gbif", ]) == 0) {
        cat("\n### Note: zero records were returned from GBIF\n")
        check_name <- TRUE
      }
      if (nrow(observations[observations$prov == "inat", ]) == 0) {
        cat("\n### Note: zero records were returned from iNaturalist\n")
        check_name <- TRUE
      }
      if (check_name) {
        cat("Please check to see if _", genus_name, " ", species_name, 
            "_ is the name used for the data provider.\n")
      }
    } else {
      cat("\n## No records returned for _", genus_name, " ", 
          species_name, "_.\n")
      cat("Check to be sure names are spelled correctly and iNat and GBIF servers are responsive")
      continue_knitting <- FALSE
    }
  }
  
  # Add genus and species columns to the data
  observations$genus_name <- genus_name
  observations$species_name <- species_name
  observations$category <- category
  observations$binomial <- paste0(genus_name, " ", species_name)
  
  if (is.null(all_obs)){
    all_obs <- observations
  } else {
    all_obs <- bind_rows(observations, 
                         all_obs)
  }
  rm(observations)
}
```

```{r status-check-data}
# If there are problems at this point, don't bother evaluating the rest of the 
# document
if (!continue_knitting) {
  knitr::knit_exit()
}
```

Retrieval from iNaturalist and GBIF returned `r nrow(all_obs)` total observations.

```{r data-cleaning, results = "asis"}
# Restrict to the days and years of interest
all_obs <- all_obs %>%
  filter(!is.na(date),
         yday(x = date) >= globals$min_julian,
         yday(x = date) <= globals$max_julian,
         year(x = date) >= globals$min_year,
         year(x = date) <= globals$max_year)

if (nrow(all_obs) == 0) {
  cat("\n## Zero observations remain after date filtering.\n")
  cat("Current filters only include ", globals$min_year, "-", globals$max_year)
  cat(" and julian days ", globals$min_julian, "-", globals$max_julian, ".\n")
  continue_knitting <- FALSE
}
```

```{r status-check-cleaning}
# If there are problems at this point, don't bother evaluating the rest of the 
# document
if (!continue_knitting) {
  knitr::knit_exit()
}
```

```{r meet-minimums, results = "asis"}
# Calculate which years meet the minimum number of required samples
year_counts <- all_obs %>%  
  group_by(year = year(date)) %>%
  summarize(obs_count = n()) %>%
  filter(obs_count >= globals$minimum_required)

# Filter obserations to only include those years with enough samples
all_obs <- all_obs %>%
  filter(year(date) %in% year_counts$year)

if (nrow(all_obs) == 0) {
  cat("\n## Zero observations remain after removing years with too few observations\n")
  cat("Current filter requires at least ", globals$minimum_required, 
      " observations per year", ".\n")
  continue_knitting <- FALSE
}
```

```{r status-check-minimums}
# If there are problems at this point, don't bother evaluating the rest of the 
# document
if (!continue_knitting) {
  knitr::knit_exit()
}
```

Following date and sample size filtering, observation counts for species:

```{r report-final-data, results = "asis"}
for (i in 1:nrow(species)) {
binomial <- paste0(species$genus[i], " ", species$species[i])
  cat("+ _", binomial,
      "_: ", nrow(all_obs[all_obs$binomial == binomial, ]),
      " observations\n", sep = "")
}
```

Observations for `r min(year(all_obs$date))` - `r max(year(all_obs$date))` (total of `r nrow(all_obs)` observations following date and sample size filtering):

```{r plot-data, include = FALSE}
# include = FALSE to prevent ggmap::get_map messages from showing up in output
# Plot a map of observations post-filtering

map_genus_name <- species$genus[species$category == "insect"][1]
map_species_name <- species$species[species$category == "insect"][1]

# Only download map files if we haven't already
map_file <- data_file <- paste0("data/maps/", map_genus_name, 
                                "-", map_species_name, "-and-hosts-map.RData")

if (file.exists(map_file)) {
  load(file = map_file)
} else {
  # Get a map image
  # Have to provide min/max for lat and long, otherwise will assume Google map,
  # regardless of value sent to source parameter
  map_bounds <- c(floor(min(all_obs$longitude)),
                floor(min(all_obs$latitude)),
                ceiling(max(all_obs$longitude)),
                ceiling(max(all_obs$latitude)))

  obs_map <- get_map(location = map_bounds,
                     source = "stamen",
                     maptype = "terrain")
  save(obs_map, file = map_file)
}

map_plot <- ggmap(ggmap = obs_map) +
  geom_point(data = all_obs,
             mapping = aes(x = longitude, y = latitude),
             size = 0.7) +
  theme_bw() +
  ylab(label = "Latitude") +
  xlab(label = "Longitude") +
  facet_wrap(~binomial)

# Plot observations by year
obs_plot <- ggplot(data = all_obs,
                   mapping = aes(x = year(date),
                                 y = yday(date),
                                 color = latitude)) +
  geom_point() +
  xlab(label = "Year") +
  scale_color_gradient(name = "Latitude") +
  ylab(label = "Day") +
  theme_bw() +
  theme(legend.position = "none", #c(0.12, 0.8),
        legend.background = element_rect(size = 1.5),
        legend.title = element_text(size = 6),
        legend.text = element_text(size = 6),
        legend.key.size = unit(x = 0.7, units = "line"),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~binomial)

```

```{r draw-observation-plots, fig.height = 5}
# Two-panel plot of map and time
suppressMessages(expr = grid.arrange(map_plot,
                                     obs_plot,
                                     nrow = 2))
```

Note in lower plot, **darker** points are from lower latitudes, while **lighter** points are from higher latitudes.

## Test for change in peak over time

Ultimately, we want to know if the changes in insect phenology are matching the changes in host phenology. That is, is the effect of year on day of year for observations the same for insect and host(s). In this case, there are two host plants, and one question is: do we analyze them separately, or do we combine them? The latter is easier to do, and, in some ways, easier to interpret:

$$
Julian\ day = \beta_0 + \beta_1Year + \beta_2Latitude + \beta_3Category \\
+ \beta_4{Year \times Category}
$$

where there are two levels to $Category$: insect and host

```{r two-species-model}
# First create year and yday variables
all_obs$yday <- yday(x = all_obs$date)
all_obs$year <- year(x = all_obs$date)

# Make the insect the reference category
all_obs$category <- factor(x = all_obs$category,
                           levels = c("insect", "host"))

# Run the model, including category and a year by category interaction term
two_sp_model <- lm(formula = yday ~ year + latitude + category + year * category,
                   data = all_obs)
two_sp_summary <- summary(two_sp_model)

#TODO: report out results of the model
```

## Model visualization
We can plot these in a couple of ways. First, we'll plot the predicted lines for several latitudes, with separate lines for insect and host. To calculate the slopes and intercepts, we need to look at the model formula and do some rearrangements. For the insect lines, the "insect" category is the reference, so we can drop all terms from the model that include $Category$ (because $Category = 0$ for all insect observations), and the model becomes:

$$
Julian\ day = \beta_0 + \beta_1Year + \beta_2Latitude
$$

And rearranging:
$$
Julian\ day = \beta_0 + \beta_2Latitude + \beta_1Year
$$

For a single value of $Latitude$, the first two terms constitute the intercept ($\beta_0 + \beta_2Latitude$) and the last coefficient is the slope ($\beta_1$).

For the host lines, $Category = 1$, so the model becomes:
$$
Julian\ day = \beta_0 + \beta_1Year + \beta_2Latitude + \beta_3 + \beta_4Year
$$

Rearranging:
$$
Julian\ day = \beta_0 + \beta_3 + \beta_2Latitude + (\beta_1 + \beta_4)Year
$$

For a single value of $Latitude$, the intercept is defined by the sum of the first three terms ($\beta_0 + \beta_3 + \beta_2Latitude$) and the slope is the sum of the coefficients $\beta_1$ and $\beta_4$.

Using those formulae, we can draw regression lines for a series of latitudes for insect and host responses through time.

```{r model-visualization}
# Extract model coefficients
two_sp_coeff <- two_sp_summary$coefficients

# Making variables to avoid confusion due to different indexing systems
b0 <- two_sp_coeff[1]
b1 <- two_sp_coeff[2]
b2 <- two_sp_coeff[3]
b3 <- two_sp_coeff[4]
b4 <- two_sp_coeff[5]

# Want separate lines for five latitudes (restricting to insect limit)
pred_lats <- quantile(x = all_obs$latitude[all_obs$category == "insect"], 
                      probs = seq(0, 1, 0.25))

# Insect model:
# Intercept = B0 + B2 * latitude
# Slope = B1

insect_intercepts <- b0 + (b2 * pred_lats)
insect_slopes <- rep(x = b1, times = length(pred_lats))

# Host model:
# Intercept = B0 + B3 + B2 * latitude
# Slope = B1 + B4

host_intercepts <- b0 + b3 + (b2 * pred_lats)
host_slopes <- rep(x = (b1 + b4), times = length(pred_lats))

ablines <- data.frame(latitude = c(pred_lats, pred_lats),
                      intercept = c(insect_intercepts, host_intercepts),
                      slope = c(insect_slopes, host_slopes),
                      category = c(rep(x = "insect", times = length(pred_lats)),
                                   rep(x = "host", times = length(pred_lats))))


lat_plot <- ggplot(data = all_obs, mapping = aes(x = year, 
                                             y = yday)) +
  geom_blank() + # Need to include or lines won't be plotted
  ylab(label = "Julian Day") +
  xlab(label = "Year") +
  geom_abline(data = ablines, mapping = aes(slope = slope, 
                                            intercept = intercept, 
                                            color = latitude,
                                            group = category,
                                            lty = category)) +
  scale_color_gradient() +
  theme_bw()

print(lat_plot)
```