---
title: "Pieris virginiensis progress report 03"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
geometry: margin=0.75in
params: 
    genus_name: "Pieris"
    species_name: "virginiensis"
    min_lon: -99
    max_lon: -45
    min_lat: 15
    max_lat: 70
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.height = 3)
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(parallel))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(ggmap))
```


```{r data-entry}
genus_name <- params$genus_name
species_name <- params$species_name
long_limits <- c(params$min_lon, params$max_lon)
lat_limits <- c(params$min_lat, params$max_lat)

# Path to folder that contains retrieval script, get_clean_obs_function.R
retrieval_path <- "../insect_migration/scripts/"
retrieval_name <- "get_clean_obs_function.R"
```

```{r check-values, results = "asis", warning = FALSE}
# Keep track of whether a report can be assembled after the setup-check
continue_knitting <- TRUE

# Seems excessive...
if (file.exists("scripts/global_values.R")) {
  source(file = "scripts/global_values.R")
  globals <- global_values()
} else {
  cat("\n## Missing global values script\n")
  cat("Make sure scripts/global_values.R exists")
  continue_knitting <- FALSE
}

# Make sure retrieval_path has trailing /
if (substr(x = retrieval_path, start = nchar(retrieval_path), stop = nchar(retrieval_path)) != "/") {
  retrieval_path <- paste0(retrieval_path, "/")
}

# Check to make sure get_clean_obs_function exists
if (!file.exists(paste0(retrieval_path, retrieval_name))) {
  cat("\n\n## Missing required function `get_clean_obs_function`\n")
  cat("Please check `retrieval_path` setting in `data-entry` chunk of Rmarkdown file. The function can be obtained from GitHub via `git clone https://github.com/keatonwilson/insect_migration.git`\n")
  continue_knitting <- FALSE
}

if ("FIXME" %in% c(species_name, genus_name)) {
  cat("\n## `FIXME` values need to be changed in `data-entry` chunk of Rmarkdown file.\n")
  continue_knitting <- FALSE
}

if (any(is.na(c(long_limits, lat_limits)))) {
  cat("\n## `NA` values need to be changed in `data-entry` chunk of Rmarkdown file.\n")
  continue_knitting <- FALSE
}

required_pkgs <- c("tidyverse", "spocc", "mapr")
missing_pkgs <- c()
for (pkg in required_pkgs) {
  if (!suppressPackageStartupMessages(library(package = pkg, logical.return = TRUE, character.only = TRUE))) {
    missing_pkgs <- c(missing_pkgs, pkg)
  }
}
if (length(missing_pkgs) > 0) { 
  cat("\n## Missing one or more required libraries.\n")
  cat("Please install the following package(s): ", paste(missing_pkgs, collapse = ", "), ".\n", sep = "")
  continue_knitting <- FALSE
}
```

```{r status-check-setup}
# If there are problems at this point, don't bother evaluating the rest of the 
# document
if (!continue_knitting) {
  knitr::knit_exit()
}
```

## Data for _`r paste(genus_name, species_name, sep = " ")`_

```{r download-data, include = FALSE}
# Use include = FALSE to suppress warning messages
source(file = paste0(retrieval_path, retrieval_name))

data_file <- paste0("data/", genus_name, "-", species_name, "-spocc.csv")

# Load in data, if they exist on disk. If not, download via spocc
if (file.exists(data_file)) {
  observations <- read_csv(file = data_file)
} else {
  observations <- get_clean_obs(genus = genus_name, 
                                species = species_name, 
                                lonlim = long_limits, 
                                latlim = lat_limits)
}

# Check for zero observations retrieved
if (nrow(observations) > 0) {
  write_csv(x = observations, path = data_file)
  check_name <- FALSE
  if (nrow(observations[observations$prov == "gbif", ]) == 0) {
    cat("\n### Note: zero records were returned from GBIF\n")
    check_name <- TRUE
  }
  if (nrow(observations[observations$prov == "inat", ]) == 0) {
    cat("\n### Note: zero records were returned from iNaturalist\n")
    check_name <- TRUE
  }
  if (check_name) {
    cat("Please check to see if _", genus_name, " ", species_name, 
        "_ is the name used for the data provider.\n")
  }
} else {
  cat("\n## No records returned for _", genus_name, " ", species_name, "_.\n")
  cat("Check to be sure names are spelled correctly and iNat and GBIF servers are responsive")
  continue_knitting <- FALSE
}
```

```{r status-check-data}
# If there are problems at this point, don't bother evaluating the rest of the 
# document
if (!continue_knitting) {
  knitr::knit_exit()
}
```

Retrieval from iNaturalist and GBIF returned `r nrow(observations)` observations.

```{r data-cleaning}
# Restrict to the days and years of interest
observations <- observations %>%
  filter(!is.na(date),
         yday(x = date) >= globals$min_julian,
         yday(x = date) <= globals$max_julian,
         year(x = date) >= globals$min_year,
         year(x = date) <= globals$max_year)

if (nrow(observations) == 0) {
  cat("\n## Zero observations remain after date filtering.\n")
  cat("Current filters only include ", globals$min_year, "-", globals$max_year)
  cat(" and julian days ", globals$min_julian, "-", globals$max_julian, ".\n")
  continue_knitting <- FALSE
}
```

```{r status-check-cleaning}
# If there are problems at this point, don't bother evaluating the rest of the 
# document
if (!continue_knitting) {
  knitr::knit_exit()
}
```

```{r meet-minimums}
# Calculate which years meet the minimum number of required samples
year_counts <- observations %>%  
  group_by(year = year(date)) %>%
  summarize(obs_count = n()) %>%
  filter(obs_count >= globals$minimum_required)

# Filter obserations to only include those years with enough samples
observations <- observations %>%
  filter(year(date) %in% year_counts$year)

if (nrow(observations) == 0) {
  cat("\n## Zero observations remain after removing years with too few observations\n")
  cat("Current filter requires at least ", globals$minimum_required, 
      " observations per year", ".\n")
  continue_knitting <- FALSE
}
```

```{r status-check-minimums}
# If there are problems at this point, don't bother evaluating the rest of the 
# document
if (!continue_knitting) {
  knitr::knit_exit()
}
```


Distribution of _`r paste(genus_name, species_name, sep = " ")`_ data for `r min(year(observations$date))` - `r max(year(observations$date))` (total of `r nrow(observations)` observations following date and sample size filtering):

```{r plot-data, include = FALSE, fig.height = 2.5}
# include = FALSE to prevent ggmap::get_map messages from showing up in output
# Plot a map of observations post-filtering

# Only download map files if we haven't already
map_file <- data_file <- paste0("data/maps/", genus_name, "-", species_name, "-map.RData")

if (file.exists(map_file)) {
  load(file = map_file)
} else {
  # Get a map image
  # Have to provide min/max for lat and long, otherwise will assume Google map,
  # regardless of value sent to source parameter
  map_bounds <- c(floor(min(observations$longitude)),
                floor(min(observations$latitude)),
                ceiling(max(observations$longitude)),
                ceiling(max(observations$latitude)))

  obs_map <- get_map(location = map_bounds,
                     source = "stamen",
                     maptype = "terrain")
  save(obs_map, file = map_file)
}

map_plot <- ggmap(ggmap = obs_map) +
  geom_point(data = observations,
             mapping = aes(x = longitude, y = latitude),
             size = 1.0) +
  theme_bw() +
  ylab(label = "Latitude") +
  xlab(label = "Longitude")

# Plot observations by year
obs_plot <- ggplot(data = observations,
                   mapping = aes(x = year(date),
                                 y = yday(date),
                                 color = latitude)) +
  geom_point() +
  xlab(label = "Year") +
  scale_color_gradient(name = "Latitude") +
  ylab(label = "Day") +
  theme_bw() +
  theme(legend.position = "none", #c(0.12, 0.8),
        legend.background = element_rect(size = 1.5),
        legend.title = element_text(size = 6),
        legend.text = element_text(size = 6),
        legend.key.size = unit(x = 0.7, units = "line"))

```

```{r draw-observation-plots}
# Two-panel plot of map and time
suppressMessages(expr = grid.arrange(map_plot, 
                                     obs_plot, 
                                     nrow = 1))
```

Note in left-hand plot, **darker** points are from lower latitudes, while **lighter** points are from higher latitudes.

# Test for change in emergence

Starting with simple linear regression, not considering variation in sampling across years (but should probably use weighted least-squares regression for this).

```{r flights-latitude}
obs <- observations %>%
  mutate(yday = yday(date),
         year = year(date)) %>%
  select(latitude, yday, year)

lat_model <- lm(yday ~ year + latitude + year*latitude, data = obs)
lat_model_coeffs <- summary(lat_model)$coefficients

model_intercept <- lat_model_coeffs[1]
model_year_slope <- lat_model_coeffs[2]
model_latitude_slope <- lat_model_coeffs[3]
model_interaction_slope <- lat_model_coeffs[4]

# To create regression lines for this plot, re-arrange the linear regression model
# yday = B0 + B1 * year + B2 * latitude + B3 + year * latitude
# yday = B0 + B2 * latitude + (B1 + B3 * latitude) * year
# where the sum of the first two terms is the intercept and (B1 + B3 * latitude)
# is the slope
pred_lats <- quantile(x = obs$latitude, probs = seq(0, 1, 0.25))
pred_intercepts <- model_intercept + model_latitude_slope * pred_lats
pred_slopes <- model_year_slope + (model_interaction_slope * pred_lats)
ablines <- data.frame(latitude = pred_lats,
                      intercept = pred_intercepts,
                      slope = pred_slopes)

lat_plot <- ggplot(data = obs, mapping = aes(x = year, 
                                             y = yday, 
                                             color = latitude)) +
  geom_point() +
  ylab(label = "Julian Day") +
  xlab(label = "Year") +
  geom_abline(data = ablines, mapping = aes(slope = slope, 
                                            intercept = intercept, 
                                            color = latitude)) +
  scale_color_gradient() +
  theme_bw()

print(lat_plot)
```


The next thing to addess is how to assess if the host plant(s) of species X are going to be available at the right latitude at the right time. Could create a vector of latitudes over which the plant and the bug occur. For those latitudes, use linear regression models to predict the yday of peak occurrance for each year. These become lines for each latitude that we can compare?

Alternatively, could compare a simple linear regression model, based on observations of insect and host plant(s) combined, with a mixed-model on the same data, where species (insect vs. host) is a random effect (need to include year as random slope). See 

+ [https://stackoverflow.com/questions/24019807/how-to-compare-a-model-with-no-random-effects-to-a-model-with-a-random-effect-us](https://stackoverflow.com/questions/24019807/how-to-compare-a-model-with-no-random-effects-to-a-model-with-a-random-effect-us).
+ [https://stats.stackexchange.com/questions/141746/comparing-mixed-effects-and-fixed-effects-models-testing-significance-of-random](https://stats.stackexchange.com/questions/141746/comparing-mixed-effects-and-fixed-effects-models-testing-significance-of-random)

This may not work for these data. From [Harrison et al. 2018](http://dx.doi.org/10.7717/peerj.4794):

>_Random effect models have several desirable properties (see above), but their use comes with some caveats. First, they are quite 'data hungry'; requiring at least five 'levels' (groups) for a random intercept term to achieve robust estimates of variance (Gelman & Hill, 2007; Harrison, 2015). With <5 levels, the mixed model may not be able to estimate the among-population variance accurately. In this case, the variance estimate will either collapse to zero, making the model equivalent to an ordinary GLM (Gelman & Hill, 2007, p. 275) or be non-zero but incorrect if the small number of groups that were sampled are not representative of true distribution of means (Harrison, 2015). Second, models can be unstable if sample sizes across groups are highly unbalanced i.e. if some groups contain very few data. These issues are especially relevant to random slope models (Grueber et al., 2011)._
