---
title: "Pieris virginiensis progress report 04"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
geometry: margin=0.75in
params: 
    min_lon: -99
    max_lon: -45
    min_lat: 15
    max_lat: 70
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.height = 3)
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(parallel))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(ggmap))
```


```{r data-entry}
# Will want to create two data frames: one with insect and one with host(s) 
species <- data.frame(genus = c("Pieris", "Cardamine", "Cardamine"),
                      species = c("virginiensis", "concatenata", "diphylla"),
                      category = c("insect", "host", "host"),
                      stringsAsFactors = FALSE)

long_limits <- c(params$min_lon, params$max_lon)
lat_limits <- c(params$min_lat, params$max_lat)

# Path to folder that contains retrieval script, get_clean_obs_function.R
retrieval_path <- "../insect_migration/scripts/"
retrieval_name <- "get_clean_obs_function.R"
```

```{r check-values, results = "asis", warning = FALSE}
# Keep track of whether a report can be assembled after the setup-check
continue_knitting <- TRUE

# Seems excessive...
if (file.exists("scripts/global_values.R")) {
  source(file = "scripts/global_values.R")
  globals <- global_values()
} else {
  cat("\n## Missing global values script\n")
  cat("Make sure scripts/global_values.R exists")
  continue_knitting <- FALSE
}

# Make sure retrieval_path has trailing /
if (substr(x = retrieval_path, start = nchar(retrieval_path), stop = nchar(retrieval_path)) != "/") {
  retrieval_path <- paste0(retrieval_path, "/")
}

# Check to make sure get_clean_obs_function exists
if (!file.exists(paste0(retrieval_path, retrieval_name))) {
  cat("\n\n## Missing required function `get_clean_obs_function`\n")
  cat("Please check `retrieval_path` setting in `data-entry` chunk of Rmarkdown file. The function can be obtained from GitHub via `git clone https://github.com/keatonwilson/insect_migration.git`\n")
  continue_knitting <- FALSE
}

if (any(is.na(c(long_limits, lat_limits)))) {
  cat("\n## `NA` values need to be changed in `data-entry` chunk of Rmarkdown file.\n")
  continue_knitting <- FALSE
}

required_pkgs <- c("tidyverse", "spocc", "mapr")
missing_pkgs <- c()
for (pkg in required_pkgs) {
  if (!suppressPackageStartupMessages(library(package = pkg, logical.return = TRUE, character.only = TRUE))) {
    missing_pkgs <- c(missing_pkgs, pkg)
  }
}
if (length(missing_pkgs) > 0) { 
  cat("\n## Missing one or more required libraries.\n")
  cat("Please install the following package(s): ", paste(missing_pkgs, collapse = ", "), ".\n", sep = "")
  continue_knitting <- FALSE
}
```

```{r status-check-setup}
# If there are problems at this point, don't bother evaluating the rest of the 
# document
if (!continue_knitting) {
  knitr::knit_exit()
}
```

```{r download-data, results = "asis", include = FALSE}
# Use include = FALSE to suppress warning messages
source(file = paste0(retrieval_path, retrieval_name))

all_obs <- NULL

for (i in 1:nrow(species)) {
  genus_name <- species$genus[i]
  species_name <- species$species[i]
  category <- species$category[i]
  
  data_file <- paste0("data/", genus_name, "-", species_name, "-spocc.csv")
  
  # Load in data, if they exist on disk. If not, download via spocc
  if (file.exists(data_file)) {
    observations <- read_csv(file = data_file)
  } else {
    observations <- get_clean_obs(genus = genus_name, 
                                  species = species_name, 
                                  lonlim = long_limits, 
                                  latlim = lat_limits)
    
    # Check for zero observations retrieved
    if (nrow(observations) > 0) {
      write_csv(x = observations, path = data_file)
      check_name <- FALSE
      if (nrow(observations[observations$prov == "gbif", ]) == 0) {
        cat("\n### Note: zero records were returned from GBIF\n")
        check_name <- TRUE
      }
      if (nrow(observations[observations$prov == "inat", ]) == 0) {
        cat("\n### Note: zero records were returned from iNaturalist\n")
        check_name <- TRUE
      }
      if (check_name) {
        cat("Please check to see if _", genus_name, " ", species_name, 
            "_ is the name used for the data provider.\n")
      }
    } else {
      cat("\n## No records returned for _", genus_name, " ", 
          species_name, "_.\n")
      cat("Check to be sure names are spelled correctly and iNat and GBIF servers are responsive")
      continue_knitting <- FALSE
    }
  }
  
  # Restrict to the days and years of interest
  observations <- observations %>%
    filter(!is.na(date),
           yday(x = date) >= globals$min_julian,
           yday(x = date) <= globals$max_julian,
           year(x = date) >= globals$min_year,
           year(x = date) <= globals$max_year)
  
  # Check for at least some observations
  if (nrow(observations) == 0) {
    cat("\n## Zero observations remain after date filtering for _", genus_name, 
        " ", species_name, "_.\n")
    cat("Current filters only include ", globals$min_year, "-", globals$max_year)
    cat(" and julian days ", globals$min_julian, "-", globals$max_julian, ".\n")
    continue_knitting <- FALSE
  }
  
  # Filter observations to only keep those from years with at least a minimum 
  # number of observations
  year_counts <- observations %>%  
    group_by(year = year(date)) %>%
    summarize(obs_count = n()) %>%
    filter(obs_count >= globals$minimum_required)
  
  # Filter obserations to only include those years with enough samples
  observations <- observations %>%
    filter(year(date) %in% year_counts$year)
  
  # Check for at least some observations after removing years with too few 
  # observations
  if (nrow(observations) == 0) {
    cat("\n## Zero observations remain after removing years with too few 
        observations for _", genus_name, " ", species_name, "_.\n")
    cat("Current filter requires at least ", globals$minimum_required, 
        " observations per year", ".\n")
    continue_knitting <- FALSE
  }
  
  # Add genus and species columns to the data
  observations$genus_name <- genus_name
  observations$species_name <- species_name
  observations$category <- category
  observations$binomial <- paste0(genus_name, " ", species_name)
    
  if (is.null(all_obs)){
    all_obs <- observations
  } else {
    all_obs <- bind_rows(observations, 
                         all_obs)
  }
  rm(observations)
}
```

```{r status-check-data}
# If there are problems at this point, don't bother evaluating the rest of the 
# document
if (!continue_knitting) {
  knitr::knit_exit()
}
```

Retrieval from iNaturalist and GBIF returned `r nrow(all_obs)` total observations.

```{r status-check-cleaning}
# If there are problems at this point, don't bother evaluating the rest of the 
# document
if (!continue_knitting) {
  knitr::knit_exit()
}
```

Following date and sample size filtering, observation counts for species:

```{r report-final-data, results = "asis"}
for (i in 1:nrow(species)) {
binomial <- paste0(species$genus[i], " ", species$species[i])
  cat("+ _", binomial,
      "_: ", nrow(all_obs[all_obs$binomial == binomial, ]),
      " observations\n", sep = "")
}
```

Observations for `r min(year(all_obs$date))` - `r max(year(all_obs$date))` (total of `r nrow(all_obs)` observations following date and sample size filtering):

```{r plot-data, include = FALSE}
# include = FALSE to prevent ggmap::get_map messages from showing up in output
# Plot a map of observations post-filtering

map_genus_name <- species$genus[species$category == "insect"][1]
map_species_name <- species$species[species$category == "insect"][1]

# Only download map files if we haven't already
map_file <- data_file <- paste0("data/maps/", map_genus_name, 
                                "-", map_species_name, "-and-hosts-map.RData")

if (file.exists(map_file)) {
  load(file = map_file)
} else {
  # Get a map image
  # Have to provide min/max for lat and long, otherwise will assume Google map,
  # regardless of value sent to source parameter
  map_bounds <- c(floor(min(all_obs$longitude)),
                floor(min(all_obs$latitude)),
                ceiling(max(all_obs$longitude)),
                ceiling(max(all_obs$latitude)))

  obs_map <- get_map(location = map_bounds,
                     source = "stamen",
                     maptype = "terrain")
  save(obs_map, file = map_file)
}

map_plot <- ggmap(ggmap = obs_map) +
  geom_point(data = all_obs,
             mapping = aes(x = longitude, y = latitude),
             size = 0.7) +
  theme_bw() +
  ylab(label = "Latitude") +
  xlab(label = "Longitude") +
  facet_wrap(~binomial)

# Plot observations by year
obs_plot <- ggplot(data = all_obs,
                   mapping = aes(x = year(date),
                                 y = yday(date),
                                 color = latitude)) +
  geom_point() +
  xlab(label = "Year") +
  scale_color_gradient(name = "Latitude") +
  ylab(label = "Day") +
  theme_bw() +
  theme(legend.position = "none", #c(0.12, 0.8),
        legend.background = element_rect(size = 1.5),
        legend.title = element_text(size = 6),
        legend.text = element_text(size = 6),
        legend.key.size = unit(x = 0.7, units = "line"),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~binomial)

```

```{r draw-observation-plots, fig.height = 5}
# Two-panel plot of map and time
suppressMessages(expr = grid.arrange(map_plot,
                                     obs_plot,
                                     nrow = 2))
```

Note in lower plot, **darker** points are from lower latitudes, while **lighter** points are from higher latitudes.

## Test for change in peak over time

Ultimately, we want to know if the changes in insect phenology are matching the changes in host phenology. That is, is the effect of year on day of year for observations the same for insect and host(s). In this case, there are two host plants, and one question is: do we analyze them separately, or do we combine them? The latter is easier to do, and, in some ways, easier to interpret:

$$
Julian\ day = \beta_0 + \beta_1Year + \beta_2Latitude + \beta_3Category \\
+ \beta_4{Year \times Category}
$$

where there are two levels to $Category$: insect and host

```{r two-species-model}
# First create year and yday variables
all_obs$yday <- yday(x = all_obs$date)
all_obs$year <- year(x = all_obs$date)

# Make the insect the reference category
all_obs$category <- factor(x = all_obs$category,
                           levels = c("insect", "host"))

# Run the model, including category and a year by category interaction term
two_sp_model <- lm(formula = yday ~ year + latitude + category + year * category,
                   data = all_obs)
two_sp_summary <- summary(two_sp_model)

#TODO: report out results of the model
```

## Model visualization
We can plot these in a couple of ways. First, we'll plot the predicted lines for several latitudes, with separate lines for insect and host. To calculate the slopes and intercepts, we need to look at the model formula and do some rearrangements. For the insect lines, the "insect" category is the reference, so we can drop all terms from the model that include $Category$ (because $Category = 0$ for all insect observations), and the model becomes:

$$
Julian\ day = \beta_0 + \beta_1Year + \beta_2Latitude
$$

And rearranging:
$$
Julian\ day = \beta_0 + \beta_2Latitude + \beta_1Year
$$

For a single value of $Latitude$, the first two terms constitute the intercept ($\beta_0 + \beta_2Latitude$) and the last coefficient is the slope ($\beta_1$).

For the host lines, $Category = 1$, so the model becomes:
$$
Julian\ day = \beta_0 + \beta_1Year + \beta_2Latitude + \beta_3 + \beta_4Year
$$

Rearranging:
$$
Julian\ day = \beta_0 + \beta_3 + \beta_2Latitude + (\beta_1 + \beta_4)Year
$$

For a single value of $Latitude$, the intercept is defined by the sum of the first three terms ($\beta_0 + \beta_3 + \beta_2Latitude$) and the slope is the sum of the coefficients $\beta_1$ and $\beta_4$.

Using those formulae, we can draw regression lines for a series of latitudes for insect and host responses through time.

```{r two-sp-model-visualization}
# Extract model coefficients
two_sp_coeff <- two_sp_summary$coefficients

# Making variables to avoid confusion due to different indexing systems
b0 <- two_sp_coeff[1]
b1 <- two_sp_coeff[2]
b2 <- two_sp_coeff[3]
b3 <- two_sp_coeff[4]
b4 <- two_sp_coeff[5]

# Want separate lines for five latitudes (restricting to insect limit)
min_lat_insect <- min(all_obs$latitude[all_obs$category == "insect"])
med_lat_insect <- median(all_obs$latitude[all_obs$category == "insect"])
max_lat_insect <- max(all_obs$latitude[all_obs$category == "insect"])
pred_lats <- c(min_lat_insect,
               med_lat_insect,
               max_lat_insect)

# Insect model:
# Intercept = B0 + B2 * latitude
# Slope = B1

insect_intercepts <- b0 + (b2 * pred_lats)
insect_slopes <- rep(x = b1, times = length(pred_lats))

# Host model:
# Intercept = B0 + B3 + B2 * latitude
# Slope = B1 + B4

host_intercepts <- b0 + b3 + (b2 * pred_lats)
host_slopes <- rep(x = (b1 + b4), times = length(pred_lats))

ablines <- data.frame(latitude = c(pred_lats, pred_lats),
                      intercept = c(insect_intercepts, host_intercepts),
                      slope = c(insect_slopes, host_slopes),
                      category = c(rep(x = "insect", times = length(pred_lats)),
                                   rep(x = "host", times = length(pred_lats))))


lat_plot <- ggplot(data = all_obs, mapping = aes(x = year, 
                                             y = yday)) +
  geom_blank() + # Need to include or lines won't be plotted
  ylab(label = "Julian Day") +
  xlab(label = "Year") +
  geom_abline(data = ablines, mapping = aes(slope = slope, 
                                            intercept = intercept, 
                                            color = latitude,
                                            group = category,
                                            lty = category)) +
  scale_color_gradient() +
  theme_bw()

print(lat_plot)
```

## Separating hosts
There is question (in my mind at least) about the lumping together of both hosts. It may be more appropriate to replace the $Category$ predictor with $Species$, keeping _P. virginensis_ as the reference for ease of interpretation.

$$
Julian\ day = \beta_0 + \beta_1Year + \beta_2Latitude + \beta_3Species \\
+ \beta_4{Year \times Species}
$$

However, because there are **three** levels to $Species$, a more accurate representation of this model would be 

$$
Julian\ day = \beta_0 + \beta_1Year + \beta_2Latitude + \beta_3concatenata + \beta_4diphylla \\
+ \beta_5{Year \times concatenata} + \beta_6{Year \times diphylla}
$$

(so long as _P. virginensis_ is reference).

```{r species-model}
all_obs$binomial <- factor(x = all_obs$binomial,
                           levels = c("Pieris virginiensis", "Cardamine concatenata", "Cardamine diphylla"))

three_sp_model <- lm(yday ~ year + latitude + binomial + year * binomial,
                    data = all_obs)
three_sp_summary <- summary(three_sp_model)
```

The model for _P. virginensis_ remains the same:
$$
Julian\ day = \beta_0 + \beta_1Year + \beta_2Latitude
$$

And rearranging:
$$
Julian\ day = \beta_0 + \beta_2Latitude + \beta_1Year
$$

For a single value of $Latitude$, the first two terms constitute the intercept ($\beta_0 + \beta_2Latitude$) and the last coefficient is the slope ($\beta_1$).


For the host lines, there are two models, one for each species of host. For _C. concatenata_:
$$
Julian\ day = \beta_0 + \beta_1Year + \beta_2Latitude + \beta_3 + \beta_5Year
$$

Rearranging:
$$
Julian\ day = \beta_0 + \beta_3 + \beta_2Latitude + (\beta_1 + \beta_5)Year
$$

So for a single value of $Latitude$, the model for _C. concatenata_ has an intercept of $\beta_0 + \beta_3 + \beta_2Latitude$ and a slope of $\beta_1$ and $\beta_5$

Similarly, the model for _C. diphylla_ can be simplified to an intercept of $\beta_0 + \beta_4 + \beta_2Latitude$ and a slope of $\beta_1 + \beta_6$

Visualizing the model:

```{r three-sp-model-visualization}
# Extract model coefficients
three_sp_coeff <- three_sp_summary$coefficients

# Making variables to avoid confusion due to different indexing systems
b0 <- three_sp_coeff[1] # Intercept
b1 <- three_sp_coeff[2] # Year
b2 <- three_sp_coeff[3] # Latitude
b3 <- three_sp_coeff[4] # concatenata
b4 <- three_sp_coeff[5] # diphylla
b5 <- three_sp_coeff[6] # concatenata X year
b6 <- three_sp_coeff[7] # diphylla X year

# Insect model:
# Intercept = B0 + B2 * latitude
# Slope = B1

insect_intercepts <- b0 + (b2 * pred_lats)
insect_slopes <- rep(x = b1, times = length(pred_lats))

# concatenata model:
# Intercept = B0 + B3 + B2 * latitude
# Slope = B1 + B5
concatenata_intercepts <- b0 + b3 + (b2 * pred_lats)
concatenata_slopes <- rep(x = (b1 + b5), times = length(pred_lats))

# diphylla model:
# Intercept = B0 + B4 + B2 * latitude
# Slope = B1 + B6
diphylla_intercepts <- b0 + b4 + (b2 * pred_lats)
diphylla_slopes <- rep(x = (b1 + b6), times = length(pred_lats))

ablines <- data.frame(latitude = c(pred_lats, pred_lats, pred_lats),
                      intercept = c(insect_intercepts, concatenata_intercepts, diphylla_intercepts),
                      slope = c(insect_slopes, concatenata_slopes, diphylla_slopes),
                      species = c(rep(x = "P. virginiensis", times = length(pred_lats)),
                                   rep(x = "C. concatenata", times = length(pred_lats)),
                                   rep(x = "C. diphylla", times = length(pred_lats))))
# Re-level to graph the way we want
ablines$species <- factor(x = ablines$species,
                          levels = c("P. virginiensis", "C. concatenata", "C. diphylla"))

lat_plot <- ggplot(data = all_obs, mapping = aes(x = year, 
                                             y = yday)) +
  geom_blank() + # Need to include or lines won't be plotted
  ylab(label = "Julian Day") +
  xlab(label = "Year") +
  geom_abline(data = ablines, mapping = aes(slope = slope, 
                                            intercept = intercept, 
                                            color = latitude,
                                            group = species,
                                            lty = species)) +
  scale_color_gradient() +
  theme_bw()

print(lat_plot)
```

Considering the slopes, we see that controlling for latitude, observations are getting earlier, but the change is happening faster in the host plants than in _P. virginiensis_:

| Species | Change |
|:--------|:------:|
| _P. virginiensis_ | `r round(abs(b1), digits = 3)` days/year      |
| _C. concatenata_  | `r round(abs(b1 + b5), digits = 3)` days/year |
| _C. diphylla_     | `r round(abs(b1 + b6), digits = 3)` days/year |

# Separate analyses

There is some question about the sample size issues among the three taxa. Is the relationship between year and yday for _P. virginiensis_ being heavily influenced by the considerably larger sample sizes of the host plants?

## _P. virginiensis_ data alone.

Consider the model 

$$
Julian\ day = \beta_0 + \beta_1Year + \beta_2Latitude
$$

```{r insect-linear-model}
insect_model <- lm(yday ~ year + latitude,
                   data = all_obs[all_obs$genus_name == "Pieris", ])
insect_model_coeff <- summary(insect_model)$coefficients

ablines <- data.frame(latitude = pred_lats,
                      intercept = insect_model_coeff[1] + insect_model_coeff[3] * pred_lats,
                      slope = insect_model_coeff[2])

lat_plot <- ggplot(data = all_obs[all_obs$genus_name == "Pieris", ], 
                   mapping = aes(x = year, y = yday)) +
  geom_blank() + # Need to include or lines won't be plotted
  ylab(label = "Julian Day") +
  xlab(label = "Year") +
  geom_abline(data = ablines, mapping = aes(slope = slope, 
                                            intercept = intercept, 
                                            color = latitude)) +
  scale_color_gradient() +
  theme_bw()

print(lat_plot)
```

Controlling for latitude, the change in _P. virginiensis_ is `r abs(round(insect_model_coeff[2], digits = 3))` days earlier per year. This is a larger change than estimated in the model that includes the host plants (`r round(abs(b1), digits = 3)` days/year).

# Interpolation
One approach used in other phenological studies uses interpolation to estimate the day on which 10% of the observations for that year occurred (and 50% and 90%). This approach should be less influenced by sample size. For example, we can look at the bottom 10% estimate for each year for _Pieris virginiensis_:

```{r lower-10-insect}
# When computing the quantile, will lose latitude information
insect_10 <- all_obs[all_obs$genus_name == "Pieris", ] %>%
  group_by(year) %>%
  summarise(yday = quantile(x = yday, probs = 0.1, names = FALSE),
            early = TRUE)

insect_10_model <- lm(yday ~ year, data = insect_10)
insect_10_summary <- summary(insect_10_model)
insect_10_coeff <- insect_10_summary$coefficients[, 1]

insect_10_plot <- ggplot(data = insect_10, mapping = aes(x = year, y = yday)) + 
  geom_point() +
  geom_abline(intercept = insect_10_coeff[1], slope = insect_10_coeff[2]) +
  ylab(label = "Julian Day") +
  xlab(label = "Year") +
  ggtitle(label = "Pieris virginiensis, 10%") +
  theme_bw()

print(insect_10_plot)
```

We can also investigate the same statistic (the day of they year on which 10% of the observations for that year were observed) for the two host plants.

```{r lower-10-all}
# Look at all species simultaneously
all_10 <- all_obs %>%
  group_by(binomial, year) %>%
  summarise(yday = quantile(x = yday, probs = 0.1, names = FALSE),
            early = TRUE)

all_10_model <- lm(yday ~ year + binomial + year * binomial,
                   data = all_10)
all_10_summary <- summary(all_10_model)
all_10_coeff <- all_10_summary$coefficients

# Extract coefficients for plot lines
insect_intercept <- all_10_coeff[1]
concatenata_intercept <- all_10_coeff[1] + all_10_coeff[3]
diphylla_intercept <- all_10_coeff[1] + all_10_coeff[4]
insect_slope <- all_10_coeff[2]
concatenata_slope <- all_10_coeff[2] + all_10_coeff[5]
diphylla_slope <- all_10_coeff[2] + all_10_coeff[6]

# Lines for plot
all_10_ablines <- data.frame(intercept = c(insect_intercept, concatenata_intercept, diphylla_intercept),
                      slope = c(insect_slope, concatenata_slope, diphylla_slope),
                      species = c("Pieris virginiensis", "Cardamine concatenata", "Cardamine diphylla"))

all_10_plot <- ggplot(data = all_10, mapping = aes(x = year, y = yday)) +
  geom_blank() +
  geom_abline(data = all_10_ablines, mapping = aes(slope = slope,
                                                   intercept = intercept,
                                                   color = species,
                                                   group = species,
                                                   lty = species)) +
  ylab(label = "Julian Day") +
  xlab(label = "Year") +
  ggtitle(label = "All species, 10%") +
  theme_bw()

print(all_10_plot)
```

Now consider the _end_ of the season, with the 90% quantile.

```{r all-90}
# Look at all species simultaneously
all_90 <- all_obs %>%
  group_by(binomial, year) %>%
  summarise(yday = quantile(x = yday, probs = 0.9, names = FALSE),
            early = TRUE)

all_90_model <- lm(yday ~ year + binomial + year * binomial,
                   data = all_90)
all_90_summary <- summary(all_90_model)
all_90_coeff <- all_90_summary$coefficients

# Extract coefficients for plot lines
insect_intercept <- all_90_coeff[1]
concatenata_intercept <- all_90_coeff[1] + all_90_coeff[3]
diphylla_intercept <- all_90_coeff[1] + all_90_coeff[4]
insect_slope <- all_90_coeff[2]
concatenata_slope <- all_90_coeff[2] + all_90_coeff[5]
diphylla_slope <- all_90_coeff[2] + all_90_coeff[6]

# Lines for plot
all_90_ablines <- data.frame(intercept = c(insect_intercept, concatenata_intercept, diphylla_intercept),
                             slope = c(insect_slope, concatenata_slope, diphylla_slope),
                             species = c("Pieris virginiensis", "Cardamine concatenata", "Cardamine diphylla"))

all_90_plot <- ggplot(data = all_90, mapping = aes(x = year, y = yday)) +
  geom_blank() +
  geom_abline(data = all_90_ablines, mapping = aes(slope = slope,
                                                   intercept = intercept,
                                                   color = species,
                                                   group = species,
                                                   lty = species)) +
  ylab(label = "Julian Day") +
  xlab(label = "Year") +
  ggtitle(label = "All species, 90%") +
  theme_bw()

print(all_90_plot)
```

Considering both 10% and 90% on the same plot:

```{r plot-10-and-10}
all_10_ablines$q <- 10
all_90_ablines$q <- 90
both_ablines <- rbind(all_10_ablines, all_90_ablines)
all_both <- rbind(all_10, all_90)

all_both_plot <- ggplot(data = all_both, mapping = aes(x = year, y = yday)) +
  geom_blank() +
  geom_abline(data = both_ablines, mapping = aes(slope = slope,
                                                   intercept = intercept,
                                                   color = species,
                                                   group = species,
                                                   lty = species)) +
  ylab(label = "Julian Day") +
  xlab(label = "Year") +
  ggtitle(label = "All species, 10% and 90%") +
  theme_bw()

print(all_both_plot)
```

**But**, we really need to account for latitude. It's not clear this approach will account for this.