---
title: "Pieris virginiensis progress report 06"
author: "Jeff Oliver"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
geometry: margin=0.75in
params: 
    min_lon: -99
    max_lon: -45
    min_lat: 15
    max_lat: 70
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.height = 3)
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(parallel))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(ggmap))
```


```{r data-entry}
# Will want to create two data frames: one with insect and one with host(s) 
species <- data.frame(genus = c("Pieris", "Cardamine", "Cardamine"),
                      species = c("virginiensis", "concatenata", "diphylla"),
                      category = c("insect", "host", "host"),
                      stringsAsFactors = FALSE)

long_limits <- c(params$min_lon, params$max_lon)
lat_limits <- c(params$min_lat, params$max_lat)

# Path to folder that contains retrieval script, get_clean_obs_function.R
retrieval_path <- "../insect_migration/scripts/"
retrieval_name <- "get_clean_obs_function.R"
```

```{r check-values, results = "asis", warning = FALSE}
# Keep track of whether a report can be assembled after the setup-check
continue_knitting <- TRUE

# Seems excessive...
if (file.exists("scripts/global_values.R")) {
  source(file = "scripts/global_values.R")
  globals <- global_values()
} else {
  cat("\n## Missing global values script\n")
  cat("Make sure scripts/global_values.R exists")
  continue_knitting <- FALSE
}

# Make sure retrieval_path has trailing /
if (substr(x = retrieval_path, start = nchar(retrieval_path), stop = nchar(retrieval_path)) != "/") {
  retrieval_path <- paste0(retrieval_path, "/")
}

# Check to make sure get_clean_obs_function exists
if (!file.exists(paste0(retrieval_path, retrieval_name))) {
  cat("\n\n## Missing required function `get_clean_obs_function`\n")
  cat("Please check `retrieval_path` setting in `data-entry` chunk of Rmarkdown file. The function can be obtained from GitHub via `git clone https://github.com/keatonwilson/insect_migration.git`\n")
  continue_knitting <- FALSE
}

if (any(is.na(c(long_limits, lat_limits)))) {
  cat("\n## `NA` values need to be changed in `data-entry` chunk of Rmarkdown file.\n")
  continue_knitting <- FALSE
}

required_pkgs <- c("tidyverse", "spocc", "mapr")
missing_pkgs <- c()
for (pkg in required_pkgs) {
  if (!suppressPackageStartupMessages(library(package = pkg, logical.return = TRUE, character.only = TRUE))) {
    missing_pkgs <- c(missing_pkgs, pkg)
  }
}
if (length(missing_pkgs) > 0) { 
  cat("\n## Missing one or more required libraries.\n")
  cat("Please install the following package(s): ", paste(missing_pkgs, collapse = ", "), ".\n", sep = "")
  continue_knitting <- FALSE
}
```

```{r status-check-setup}
# If there are problems at this point, don't bother evaluating the rest of the 
# document
if (!continue_knitting) {
  knitr::knit_exit()
}
```

```{r download-data, results = "asis", include = FALSE}
# Use include = FALSE to suppress warning messages
source(file = paste0(retrieval_path, retrieval_name))

all_obs <- NULL

for (i in 1:nrow(species)) {
  genus_name <- species$genus[i]
  species_name <- species$species[i]
  category <- species$category[i]
  
  data_file <- paste0("data/", genus_name, "-", species_name, "-spocc.csv")
  
  # Load in data, if they exist on disk. If not, download via spocc
  if (file.exists(data_file)) {
    observations <- read_csv(file = data_file)
  } else {
    observations <- get_clean_obs(genus = genus_name, 
                                  species = species_name, 
                                  lonlim = long_limits, 
                                  latlim = lat_limits)
    
    # Check for zero observations retrieved
    if (nrow(observations) > 0) {
      write_csv(x = observations, path = data_file)
      check_name <- FALSE
      if (nrow(observations[observations$prov == "gbif", ]) == 0) {
        cat("\n### Note: zero records were returned from GBIF\n")
        check_name <- TRUE
      }
      if (nrow(observations[observations$prov == "inat", ]) == 0) {
        cat("\n### Note: zero records were returned from iNaturalist\n")
        check_name <- TRUE
      }
      if (check_name) {
        cat("Please check to see if _", genus_name, " ", species_name, 
            "_ is the name used for the data provider.\n")
      }
    } else {
      cat("\n## No records returned for _", genus_name, " ", 
          species_name, "_.\n")
      cat("Check to be sure names are spelled correctly and iNat and GBIF servers are responsive")
      continue_knitting <- FALSE
    }
  }
  
  # Restrict to the days and years of interest
  observations <- observations %>%
    filter(!is.na(date),
           yday(x = date) >= globals$min_julian,
           yday(x = date) <= globals$max_julian,
           year(x = date) >= globals$min_year,
           year(x = date) <= globals$max_year)
  
  # Check for at least some observations
  if (nrow(observations) == 0) {
    cat("\n## Zero observations remain after date filtering for _", genus_name, 
        " ", species_name, "_.\n")
    cat("Current filters only include ", globals$min_year, "-", globals$max_year)
    cat(" and julian days ", globals$min_julian, "-", globals$max_julian, ".\n")
    continue_knitting <- FALSE
  }
  
  # Filter observations to only keep those from years with at least a minimum 
  # number of observations
  year_counts <- observations %>%  
    group_by(year = year(date)) %>%
    summarize(obs_count = n()) %>%
    filter(obs_count >= globals$minimum_required)
  
  # Filter obserations to only include those years with enough samples
  observations <- observations %>%
    filter(year(date) %in% year_counts$year)
  
  # Check for at least some observations after removing years with too few 
  # observations
  if (nrow(observations) == 0) {
    cat("\n## Zero observations remain after removing years with too few 
        observations for _", genus_name, " ", species_name, "_.\n")
    cat("Current filter requires at least ", globals$minimum_required, 
        " observations per year", ".\n")
    continue_knitting <- FALSE
  }
  
  # Add genus and species columns to the data
  observations$genus_name <- genus_name
  observations$species_name <- species_name
  observations$category <- category
  observations$binomial <- paste0(genus_name, " ", species_name)
    
  if (is.null(all_obs)){
    all_obs <- observations
  } else {
    all_obs <- bind_rows(observations, 
                         all_obs)
  }
  rm(observations)
}
```

```{r status-check-data}
# If there are problems at this point, don't bother evaluating the rest of the 
# document
if (!continue_knitting) {
  knitr::knit_exit()
}
```

Retrieval from iNaturalist and GBIF returned `r nrow(all_obs)` total observations.

```{r status-check-cleaning}
# If there are problems at this point, don't bother evaluating the rest of the 
# document
if (!continue_knitting) {
  knitr::knit_exit()
}
```

Following date and sample size filtering, observation counts for species:

```{r report-final-data, results = "asis"}
for (i in 1:nrow(species)) {
binomial <- paste0(species$genus[i], " ", species$species[i])
  cat("+ _", binomial,
      "_: ", nrow(all_obs[all_obs$binomial == binomial, ]),
      " observations\n", sep = "")
}
```

Observations for `r min(year(all_obs$date))` - `r max(year(all_obs$date))` (total of `r nrow(all_obs)` observations following date and sample size filtering):

```{r plot-data, include = FALSE}
# include = FALSE to prevent ggmap::get_map messages from showing up in output
# Plot a map of observations post-filtering

map_genus_name <- species$genus[species$category == "insect"][1]
map_species_name <- species$species[species$category == "insect"][1]

# Only download map files if we haven't already
map_file <- data_file <- paste0("data/maps/", map_genus_name, 
                                "-", map_species_name, "-and-hosts-map.RData")

if (file.exists(map_file)) {
  load(file = map_file)
} else {
  # Get a map image
  # Have to provide min/max for lat and long, otherwise will assume Google map,
  # regardless of value sent to source parameter
  map_bounds <- c(floor(min(all_obs$longitude)),
                floor(min(all_obs$latitude)),
                ceiling(max(all_obs$longitude)),
                ceiling(max(all_obs$latitude)))

  obs_map <- get_map(location = map_bounds,
                     source = "stamen",
                     maptype = "terrain")
  save(obs_map, file = map_file)
}

map_plot <- ggmap(ggmap = obs_map) +
  geom_point(data = all_obs,
             mapping = aes(x = longitude, y = latitude),
             size = 0.7) +
  theme_bw() +
  ylab(label = "Latitude") +
  xlab(label = "Longitude") +
  facet_wrap(~binomial)

# Plot observations by year
obs_plot <- ggplot(data = all_obs,
                   mapping = aes(x = year(date),
                                 y = yday(date),
                                 color = latitude)) +
  geom_point() +
  xlab(label = "Year") +
  scale_color_gradient(name = "Latitude") +
  ylab(label = "Day") +
  theme_bw() +
  theme(legend.position = "none", #c(0.12, 0.8),
        legend.background = element_rect(size = 1.5),
        legend.title = element_text(size = 6),
        legend.text = element_text(size = 6),
        legend.key.size = unit(x = 0.7, units = "line"),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~binomial)

```

```{r draw-observation-plots, fig.height = 5}
# Two-panel plot of map and time
suppressMessages(expr = grid.arrange(map_plot,
                                     obs_plot,
                                     nrow = 2))
```

Note in lower plot, **darker** points are from lower latitudes, while **lighter** points are from higher latitudes.

## Models allowing species x latitude interactions

Previous models only included species x year interactions, but since these things all have different altitudinal ranges, would be useful to include that interaction as well. 

$$
Julian\ day = \beta_0 + \beta_1Year + \beta_2Latitude + \beta_3Species \\
+ \beta_4{Year \times Species} + \beta_5{Latitude \times Species}
$$

Expanding $Species$ because there are _three_ levels, this becomes:

However, because there are **three** levels to $Species$, a more accurate representation of this model would be 

$$
Julian\ day = \beta_0 + \beta_1Year + \beta_2Latitude + \beta_3concatenata + \beta_4diphylla \\
+ \beta_5{Year \times concatenata} + \beta_6{Year \times diphylla} \\
+ \beta_7{Latitude \times concatenata} + \beta_8{Latitude \times diphylla}
$$

(so long as _P. virginensis_ is reference).

```{r species-model}
# First create year and yday variables
all_obs$yday <- yday(x = all_obs$date)
all_obs$year <- year(x = all_obs$date)

# Ensure leveling has P. virginiensis as reference
all_obs$binomial <- factor(x = all_obs$binomial,
                           levels = c("Pieris virginiensis", "Cardamine concatenata", "Cardamine diphylla"))

three_sp_model <- lm(yday ~ year + latitude + binomial + year * binomial + latitude * binomial,
                    data = all_obs)
three_sp_summary <- summary(three_sp_model)
```

The model for _P. virginensis_ remains the same:
$$
Julian\ day = \beta_0 + \beta_1Year + \beta_2Latitude
$$

And rearranging:
$$
Julian\ day = \beta_0 + \beta_2Latitude + \beta_1Year
$$

For a single value of $Latitude$, the first two terms constitute the intercept ($\beta_0 + \beta_2Latitude$) and the last coefficient is the slope ($\beta_1$).

For the host lines, there are two models, one for each species of host. For _C. concatenata_:
$$
Julian\ day = \beta_0 + \beta_1Year + \beta_2Latitude + \beta_3 + \beta_5Year + \beta_7Latitude
$$

Rearranging:
$$
Julian\ day = \beta_0 + \beta_3 + (\beta_1 + \beta_5)Year + (\beta_2 + \beta_7)Latitude
$$

So for a single value of $Latitude$, the model for _C. concatenata_ has an intercept of $\beta_0 + \beta_3 + (\beta_2 + \beta_7)Latitude$ and a slope of $\beta_1$ and $\beta_5$.

Similarly, the model for _C. diphylla_ can be simplified to an intercept of $\beta_0 + \beta_4 + (\beta_2 + \beta_8)Latitude$ and a slope of $\beta_1 + \beta_6$.


```{r three-sp-model-visualization}
# Extract model coefficients
three_sp_coeff <- three_sp_summary$coefficients

# Making variables to avoid confusion due to different indexing systems
b0 <- three_sp_coeff[1] # Intercept
b1 <- three_sp_coeff[2] # Year
b2 <- three_sp_coeff[3] # Latitude
b3 <- three_sp_coeff[4] # concatenata
b4 <- three_sp_coeff[5] # diphylla
b5 <- three_sp_coeff[6] # concatenata X year
b6 <- three_sp_coeff[7] # diphylla X year
b7 <- three_sp_coeff[8] # concatenata X latitude
b8 <- three_sp_coeff[9] # diphylla X latitude

# Want separate lines for five latitudes (restricting to insect limit)
min_lat_insect <- min(all_obs$latitude[all_obs$category == "insect"])
med_lat_insect <- median(all_obs$latitude[all_obs$category == "insect"])
max_lat_insect <- max(all_obs$latitude[all_obs$category == "insect"])
pred_lats <- c(min_lat_insect,
               med_lat_insect,
               max_lat_insect)

# Insect model:
# Intercept = B0 + B2 * latitude
# Slope = B1

insect_intercepts <- b0 + (b2 * pred_lats)
insect_slopes <- rep(x = b1, times = length(pred_lats))

# concatenata model:
# Intercept = B0 + B3 + (B2 + B7) * latitude
# Slope = B1 + B5
concatenata_intercepts <- b0 + b3 + ((b2 + b7) * pred_lats)
concatenata_slopes <- rep(x = (b1 + b5), times = length(pred_lats))

# diphylla model:
# Intercept = B0 + B4 + (B2 + B8) * latitude
# Slope = B1 + B6
diphylla_intercepts <- b0 + b4 + ((b2 + b8) * pred_lats)
diphylla_slopes <- rep(x = (b1 + b6), times = length(pred_lats))

ablines <- data.frame(latitude = c(pred_lats, pred_lats, pred_lats),
                      intercept = c(insect_intercepts, concatenata_intercepts, diphylla_intercepts),
                      slope = c(insect_slopes, concatenata_slopes, diphylla_slopes),
                      species = c(rep(x = "P. virginiensis", times = length(pred_lats)),
                                   rep(x = "C. concatenata", times = length(pred_lats)),
                                   rep(x = "C. diphylla", times = length(pred_lats))))
# Re-level to graph the way we want
ablines$species <- factor(x = ablines$species,
                          levels = c("P. virginiensis", "C. concatenata", "C. diphylla"))

lat_plot <- ggplot(data = all_obs, mapping = aes(x = year, 
                                             y = yday)) +
  geom_blank() + # Need to include or lines won't be plotted
  ylab(label = "Julian Day") +
  xlab(label = "Year") +
  geom_abline(data = ablines, mapping = aes(slope = slope, 
                                            intercept = intercept, 
                                            color = latitude,
                                            group = species,
                                            lty = species)) +
  scale_color_gradient() +
  theme_bw()

print(lat_plot)
```

Considering the slopes, we see that controlling for latitude, observations are getting earlier, but the change is happening faster in the host plants than in _P. virginiensis_:

| Species | Change |
|:--------|:------:|
| _P. virginiensis_ | `r round(abs(b1), digits = 3)` days/year      |
| _C. concatenata_  | `r round(abs(b1 + b5), digits = 3)` days/year |
| _C. diphylla_     | `r round(abs(b1 + b6), digits = 3)` days/year |

